---
title: "HW-04"
author: "Sterling Hayden"
date: "2024-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(fpp3)
library(fable)
library(fabletools)
library(lubridate)
library(expsmooth)
library(lmtest)
library(zoo)
library(seasonal)
library(ggplot2)
library(seasonalview)
library(aTSA)
library(tsibble)
library(fable)
library(tidyverse)
library(imputeTS)
```


# Creating tsible data

```{r}
#read in the csv
train <- read.csv('hrl_load_metered.csv')

#convert to a tsibble
train$datetime_beginning_ept <- as.POSIXct(train$datetime_beginning_ept,
                                              format = "%m/%d/%y %H:%M",
                                              tz = "America/New_York")

head(train)
```

We have to account for multiple 2:00am’s in the spring. Lets deal with those duplicates
```{r}
#remove duplicates by grouping then calculating the average then ungrouping
train <- train %>%
  group_by(datetime_beginning_ept) %>%
  summarise(
    mw = mean(mw, na.rm = TRUE), #calculate the mean mw for each datetime_beginning_ept
    .groups = 'drop' #ungroup the data
  )
```

We now can convert it to a tsibble.
```{r}
train.ts <- train.ts %>% 
  select(datetime_beginning_ept, mw) %>% 
  as_tsibble(index = datetime_beginning_ept)

head(train.ts)
```

We also have to account for the lost 2:00am's in the fall.
```{r}
#impute the missing data with the average between the t-1 and t+1
train.ts <- train.ts %>%
  na_interpolation(option = "spline")
```
CAN SOMEONE MAKE SURE THAT THIS IS CORRECT


# Looking into the time series
```{r}
autoplot(train.ts, mw)
```





# Build an appropriate Exponential Smoothing Model. 
Forecast this model for your validation set only. 
Calculate the MAE and MAPE for the validation set. 




# Build a seasonal ARIMA model. 
Describe the approach you used to select the lags of the model.  
Forecast this model for your validation set only. 
Calculate the MAE and MAPE for the validation set. 








