---
title: "HW-02"
author: "Sterling Hayden"
date: "2024-09-04"
output: html_document
---
PJM is a regional transmission organization (RTO) that provides resources for the market of
wholesale electricity in all or parts of 13 states and the District of Columbia (source: PJM.com).
PJM, henceforth referred to as the client, would like to contract your services in understanding
metered load Megawatts for the AEP Appalachian Power transmission zone
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(fpp3)
library(fable)
library(fabletools)
library(lubridate)
library(expsmooth)
library(lmtest)
library(zoo)
library(seasonal)
library(ggplot2)
library(seasonalview)
library(aTSA)
library(tsibble)
library(fable)
library(tidyverse)
```

# Creating tsible data
For this analysis,the client would like the data to be rolled up to total daily MW and averaged by month
```{r}
eng <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")

eng_daily <- eng %>%
  mutate(date = as_date(mdy_hm(datetime_beginning_ept))) %>%
  group_by(date) %>%
  summarise(total_daily_usage = sum(mw, na.rm = TRUE))

eng_daily <- eng_daily %>% 
  mutate(monthly = (format(as.Date(date), "%m-%Y")))

eng_monthly <- eng_daily %>% 
  group_by(monthly) %>% 
  summarise(avg_monthly_usage = mean(total_daily_usage))

eng_monthly <- eng_monthly %>% 
  mutate(mon = my(monthly))

eng_monthly <- eng_monthly %>%
  mutate(mony = as.yearmon(mon))
```

The training data should be from August 2018-July 2022; the validation data set should be from
August 2022-July 2023; and the test data should be from August 2023-July 2024.
```{r}
train <- eng_monthly %>%
  filter(between(mon, as.Date('2018-08-01'), as.Date('2022-07-31')))
train$mony <- yearmonth(train$mony)
train.ts <- train %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)

val <- eng_monthly %>%
  filter(between(mon, as.Date('2022-08-01'), as.Date('2023-07-31')))
val$mony <- yearmonth(val$mony)
val.ts <- val %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)

test <- eng_monthly %>%
  filter(between(mon, as.Date('2023-08-01'), as.Date('2024-07-31')))
test$mony <- yearmonth(val$mony)
test.ts <- test %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)
```


# Creation of the best ESM model
Create a simple exp smoothing, Holt exp smoothing, and a Holt-Winters exp smoothing model

first lets take a look at the ts plot
```{r}
autoplot(train.ts, avg_monthly_usage) +
  xlab("Date by Month") +
  ylab("Mean Total Daily MW") +
  ggtitle("Mean Total Daily MW by Month From August 2018 - July 2022")
```
To me this is pretty clearly an additive model.

## Simple exponential smoothing
```{r}
# train the model
fit <- train.ts |>
  model(ETS(avg_monthly_usage ~ error("A") + trend("N") + season("N")))

# forecast the model
fc <- fit |> fabletools::forecast(h = nrow(val.ts))

# plot the forecasted data
fc |> autoplot(bind_rows(train.ts, val.ts)) + 
  geom_line(aes(y = .fitted), col="#D55E00", data = augment(fit)) +
  labs(y="Mean Total Daily MW", title="Simple Exponential Smoothing Forecast")
```


Now lets look how well we forcasted for just the validation set
```{r}
autoplot(fc) +
  autolayer(val.ts, avg_monthly_usage, color = "blue")
```


## Holt’s linear trend method

```{r}
# train the model
fit.H <- train.ts |>
  model(ETS(avg_monthly_usage ~ error("A") + trend("A") + season("N")))

# forecast the model
fc.H <- fit.H |> fabletools::forecast(h = nrow(val.ts))

autoplot(fc.H) +
  autolayer(val.ts, avg_monthly_usage, color = "blue")
```


## Damped Holt's method
```{r}
# train the model
fit.Hd <- train.ts |>
  model(ETS(avg_monthly_usage ~ error("A") + trend("Ad", phi = 0.9) + season("N")))

# forecast the model
fc.Hd <- fit.Hd |> fabletools::forecast(h = nrow(val.ts))

autoplot(fc.Hd) +
  autolayer(val.ts, avg_monthly_usage, color = "blue")
```

```{r}
train.ts |>
  model(
    `Holt's method` = ETS(avg_monthly_usage ~ error("A") +
                       trend("A") + season("N")),
    `Damped Holt's method` = ETS(avg_monthly_usage ~ error("A") +
                       trend("Ad", phi = 0.9) + season("N"))
  ) |>
  fabletools::forecast(h = nrow(val.ts)) |>
  autoplot(train.ts, level = NULL) +
  labs(title = "Australian population",
       y = "Millions") +
  guides(colour = guide_legend(title = "Forecast"))
```


## Holt-Winters’ method

```{r}
# train the model
fit.HWa <- train.ts |>
  model(ETS(avg_monthly_usage ~ error("A") + trend("A") + season("A")))

# forecast the model
fc.HWa <- fit.HWa |> fabletools::forecast(h = nrow(val.ts))

autoplot(fc.HWa) +
  autolayer(val.ts, avg_monthly_usage, color = "blue")
```


```{r}
# train the model
fit.HWm <- train.ts |>
  model(ETS(avg_monthly_usage ~ error("M") + trend("A") + season("M")))

# forecast the model
fc.HWm <- fit.HWm |> fabletools::forecast(h = nrow(val.ts))

autoplot(fc.HWm) +
  autolayer(val.ts, avg_monthly_usage, color = "blue")
```

























