---
title: "HW-00"
author: "Sterling Hayden"
date: "2024-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The usairlines data set contains monthly passenger information for the united states for the years 1990 to 2008.  Make this data set into a tsibble.  Some helpful codes:
1. Pull data into R and (call it air to use code below)
2. Use following code to create tsibble: 
```{r}
library(tidyverse)
air <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/usairlines.csv")
head(air)
```

```{r}
library(tsibble)
air.ts<-air %>% mutate(date=mdy(paste(Month, "1",Year))) %>% 
mutate(Month.ts = yearmonth(date)) %>% 
as_tsibble(index = Month.ts)

head(air.ts)
```



How many implicit missing values are in this data set?
```{r}
count_gaps(air.ts)
```


What is the lowest value you see on the y-axis (actual number value on the graph....using the tick-marks on y-axis)? 
Looking at the plot created in the last question, the series appears to be trending upward.?
```{r}
library(fable)
autoplot(air.ts, Passengers)
```
```{r}
min(air.ts$Passengers)
```



We will now switch over to an energy data set. Read in the data set energy_F2024.csv.  We will be using the date information contained in the datetime_beginning_ept and the mw (megawatt) variable.
```{r}
eng <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")
head(eng)
```
Using the energy data set, aggregate the data to be daily mw totals.  What is the largest daily mw total (rounded to the nearest mw....no decimal places)?
```{r}
eng.ts <- eng %>%
  # Convert to datetime and extract date
  mutate(date_utc = mdy_hm(datetime_beginning_ept),
         date_only = date(date_utc)) %>%
  group_by(date_only) %>%
  summarise(daily_mw_total = sum(mw)) %>%
  as_tsibble(index = date_only)


eng.ts
```

```{r}
max(eng.ts$daily_mw_total)
```

How many implicit gaps are there?
```{r}
count_gaps(eng.ts)
```