---
title: "HW-03"
author: "Sterling Hayden"
date: "2024-09-13"
output: html_document
---

PJM is a regional transmission organization (RTO) that provides resources for the market of
wholesale electricity in all or parts of 13 states and the District of Columbia (source: PJM.com).
PJM, henceforth referred to as the client, would like to contract your services in understanding
metered load Megawatts for the AEP Appalachian Power transmission zone. For this analysis,
the client would like the data to be rolled up to total daily MW and averaged by month. The
training data should be from August 2018-July 2022; the validation data set should be from
August 2022-July 2023; and the test data should be from August 2023-July 2024
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(fpp3)
library(fable)
library(fabletools)
library(lubridate)
library(expsmooth)
library(lmtest)
library(zoo)
library(seasonal)
library(ggplot2)
library(seasonalview)
library(aTSA)
library(tsibble)
library(fable)
library(tidyverse)
```


# Creating tsible data
For this analysis,the client would like the data to be rolled up to total daily MW and averaged by month
```{r}
eng <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")

eng_daily <- eng %>%
  mutate(date = as_date(mdy_hm(datetime_beginning_ept))) %>%
  group_by(date) %>%
  summarise(total_daily_usage = sum(mw, na.rm = TRUE))

eng_daily <- eng_daily %>% 
  mutate(monthly = (format(as.Date(date), "%m-%Y")))

eng_monthly <- eng_daily %>% 
  group_by(monthly) %>% 
  summarise(avg_monthly_usage = mean(total_daily_usage))

eng_monthly <- eng_monthly %>% 
  mutate(mon = my(monthly))

eng_monthly <- eng_monthly %>%
  mutate(mony = as.yearmon(mon))
```

The training data should be from August 2018-July 2022; the validation data set should be from
August 2022-July 2023; and the test data should be from August 2023-July 2024.
```{r}
train <- eng_monthly %>%
  filter(between(mon, as.Date('2018-08-01'), as.Date('2022-07-31')))
train$mony <- yearmonth(train$mony)
train.ts <- train %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)

val <- eng_monthly %>%
  filter(between(mon, as.Date('2022-08-01'), as.Date('2023-07-31')))
val$mony <- yearmonth(val$mony)
val.ts <- val %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)

test <- eng_monthly %>%
  filter(between(mon, as.Date('2023-08-01'), as.Date('2024-07-31')))
test$mony <- yearmonth(val$mony)
test.ts <- test %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)
```


# Checking Stationarity

Note: I tried the log and sqrt transformation to fix this issue but those didn't work.
```{r}
ggAcf(train.ts$avg_monthly_usage,lag=25)
```
We see seasonality in our spikes. We will have to do something about this.  


```{r}
ggPacf(train.ts$avg_monthly_usage,lag=25)
ggPacf(train.ts$avg_monthly_usage,lag=15)
```
The PACF looks much better than the ACF. We see spikes at 1,2,3,7


## What if we took a difference?

```{r}
train.ts |>
  features(avg_monthly_usage, unitroot_kpss)
```
The the unit root test says that our data is stationary and does not even need to be differenced. 


For sake of argument lets looka t what the ACF and PACF would look like if we differenced.
```{r}
ggAcf(difference(train.ts$avg_monthly_usage),lag=25)
ggPacf(difference(train.ts$avg_monthly_usage),lag=25)
```
The plots do not look any better.  I'm pretty sure the data just needs to be seasonaly differenced to fix the issue we are seeing. 






