---
title: "HW-01"
author: "Sterling Hayden"
date: "2024-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(fpp3)
library(fable)
library(fabletools)
library(lubridate)
library(expsmooth)
library(lmtest)
library(zoo)
library(seasonal)
library(ggplot2)
library(seasonalview)
library(aTSA)
library(tsibble)
library(fable)
library(tidyverse)
```

```{r}
eng <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")
head(eng)
```

```{r}
eng_daily <- eng %>%
  mutate(date = as_date(mdy_hm(datetime_beginning_ept))) %>%
  group_by(date) %>%
  summarise(total_daily_usage = sum(mw, na.rm = TRUE))

head(eng_daily)
```
```{r}
eng_daily <- eng_daily %>% 
  mutate(monthly = (format(as.Date(date), "%m-%Y")))
eng_daily

eng_monthly <- eng_daily %>% 
  group_by(monthly) %>% 
  summarise(avg_monthly_usage = mean(total_daily_usage))
eng_monthly

eng_monthly <- eng_monthly %>% 
  mutate(mon = my(monthly))
eng_monthly

eng_monthly <- eng_monthly %>%
  mutate(mony = as.yearmon(mon))
eng_monthly
```


split up into train, val, and test
Training should be from August 2018 - July 2022, validation should be from August 2022 - July 2023 and test should be from August 2023 - July 2024
```{r}
train <- eng_monthly %>%
  filter(between(mon, as.Date('2018-08-01'), as.Date('2022-07-31')))

val <- eng_monthly %>%
  filter(between(mon, as.Date('2022-08-01'), as.Date('2023-07-31')))

test <- eng_monthly %>%
  filter(between(mon, as.Date('2023-08-01'), as.Date('2024-07-31')))
```


Using the total daily average by month in the training data set, what is the median value? 
```{r}
median(train$avg_monthly_usage)
```

Decompose your training data by using the STL decomposition assuming there is a monthly seasonality.
```{r}
train$mony <- yearmonth(train$mony)
train.ts <- train %>% 
  select(mony, avg_monthly_usage) %>% 
  as_tsibble(index = mony)
```

```{r}
dcmp <- train.ts |>
  model(stl = STL(avg_monthly_usage))
components(dcmp)

components(dcmp) |> autoplot() + theme_classic()
```

Create a time plot of the total daily AVERAGE MW by month and upload it. Be sure your graph has appropriate axes, labels and title.
```{r}
autoplot(train.ts, avg_monthly_usage) +
  xlab("Date by Month") +
  ylab("Mean Total Daily MW") +
  ggtitle("Mean Total Daily MW by Month From August 2018 - July 2022")
```

What is your value for Fs, Ft?
```{r}
train.ts |>
  features(avg_monthly_usage, feat_stl)
```
F_t = strength of the trend = 0.3680337  
F_s = strength of seasonality = 0.9141054.
"A series with seasonal strength F_s close to 0 exhibits almost no seasonality, while a series with strong seasonality will have F_s close to 1" same can be said about trend.

